---
title: "Ocean Plastics Pollution"
author: "Rachna Radhakrishna & Sravan Kumar "
format: 
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
editor: visual
---

## Executive Summary

This is a report on Global plastic. Here we mainly focuses on what are the companies that are responsible for most plastic waste, the data is taken from 2021 Tidy Tuesday project, provided by Break Free From Plastic. The data includes plastic items collected during beach cleanups in over 50 countries. It records details such as which type of plastic like PET and HDPE. And the company that produced it like Coca-Cola or Nestle, and where it was found.

Here our main aim is to find out which companies are responsible for the most plastic waste and how plastic pollution differs from country to country. By doing this, we hope to understand where the biggest problems are and suggest ideas for how companies and governments can take better action against plastic pollution.

------------------------------------------------------------------------

## Introduction

Plastic pollution is a global environmental challenge affecting marine ecosystems and human health. This report uses data from Break Free From Plastic, shared via TidyTuesday, to identify top contributors to plastic waste. By analyzing company-wise and country-wise trends, we aim to uncover key patterns. The dataset includes collection information from global cleanups. The analysis spans plastic types (PET, HDPE, etc.), total items collected, and contributors. Previous studies highlight Coca-Cola, Nestlé, and Unilever as frequent polluters. This report revalidates these findings using 2021 data. It also examines changes over time. Insights can help shape policies. The ultimate goal is actionable recommendations for reducing waste.

## Methodology

Before starting the analysis, we cleaned the dataset to make it consistent and easier to work with. Company names were cleaned by fixing capital letters, removing extra spaces, and grouping unclear labels like "Null", "N/A", and "Assorted" under one category called "Unbranded". We also removed rows labeled "Grand Total" since they were just totals and not real data points. Country names were tidied up to fix spelling and formatting issues (like changing "Cote D Ivoire" to "Côte d'Ivoire"). We made sure all plastic count columns were treated as numbers, and replaced missing values with zero when it made sense.

```{r load-libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(viridis)
library(knitr)
library(ggthemes)
library(forcats)
library(readr)
library(kableExtra)
library(skimr)
library(stringr)
library(ggplot2)
```

**Loading The Data:**

```{r load-data-raw, echo=FALSE, message=FALSE, warning=FALSE}
# Load and clean data
plastics <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-01-26/plastics.csv") %>%
  clean_names()
```

**Cleaning Plastic Data Set:**

```{r clean-data, echo=FALSE, message=FALSE, warning=FALSE}
# Clean data
cleaned_plastics <- plastics %>%
  clean_names() %>%
  mutate(
    # Clean parent_company names
    parent_company = str_trim(str_to_title(parent_company)),
    parent_company = str_replace_all(parent_company, "Coca Cola", "The Coca-Cola Company"),
    parent_company = str_replace_all(parent_company, "Nestlé", "Nestle"),
    parent_company = case_when(
      str_detect(parent_company, "(?i)^(Grand Total)$") ~ NA_character_,
      str_detect(parent_company, "(?i)^(Null|N/?A|Assorted|Unbranded)$") ~ "Unbranded",
      TRUE ~ parent_company
    )
  ) %>%
  # Clean country names
  mutate(
    country = str_trim(str_to_title(country)),
    country = str_replace_all(country, "_", " "),
    country = case_when(
      country == "Empty" ~ NA_character_,
      country == "Cote D Ivoire" ~ "Côte d'Ivoire",
      country == "Cote D ivoire" ~ "Côte d'Ivoire",
      country == "Taiwan Republic Of China (Roc)" ~ "Taiwan",
      country == "Taiwan  Republic Of China (Roc)" ~ "Taiwan",
      country == "Korea" ~ "South Korea",
      country == "United Kingdom Of Great Britain & Northern Ireland" ~ "United Kingdom",
      country == "United States Of America" ~ "United States",
      TRUE ~ country
    )
  ) %>%
  filter(!is.na(parent_company), !is.na(country))
```

------------------------------------------------------------------------

### What’s in the `cleaned_plastics` dataset?

| Column           | Description                                     |
|------------------|-------------------------------------------------|
| `country`        | Country where the cleanup took place            |
| `year`           | Year of data collection (2019 or 2020)          |
| `parent_company` | Brand or company name (e.g., Coca-Cola, Nestle) |
| `empty`          | Count of unidentifiable or label-less plastics  |
| `hdpe`           | Count of HDPE (High-Density Polyethylene) items |
| `ldpe`           | Count of LDPE (Low-Density Polyethylene) items  |
| `o`              | Count of miscellaneous or other plastic types   |
| `pet`            | Count of PET (Polyethylene Terephthalate) items |
| `pp`             | Count of PP (Polypropylene) items               |
| `ps`             | Count of PS (Polystyrene) items                 |
| `pvc`            | Count of PVC (Polyvinyl Chloride) items         |
| `grand_total`    | Total plastic items collected (all types)       |
| `num_events`     | Number of cleanup events                        |
| `volunteers`     | Number of participants involved in cleanups     |

------------------------------------------------------------------------

**Most Common Plastic Types Collected:**

```{r simple_bar_plot, echo=FALSE, message=FALSE, warning=FALSE}
# Define the plastic types
plastic_types <- c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc")

# Summarise total items by type from cleaned dataset
plastic_counts <- cleaned_plastics %>%
  select(all_of(plastic_types)) %>%
  summarise(across(everything(), sum, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), names_to = "plastic_type", values_to = "total_items") %>%
  arrange(desc(total_items))

# Plot
ggplot(plastic_counts, aes(x = reorder(plastic_type, total_items), y = total_items, fill = plastic_type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d(option = "E") +  # Use colorblind-friendly palette
  labs(
    title = "Fig.1 - Most Common Plastic Types Collected",
    x = "Plastic Type",
    y = "Total Items Collected"
  ) +
  theme_light(base_size = 10)

  

```

The above bar chat indicates which type of plastic were most commonly collected during cleanup events around the world and each bar represents a specific type of plastic, and the height of the bar shows the total number of items made from that material. The chat includes well known plastics like:\
**PET** (Polyethylene Terephthalate): Commonly used in plastic drink bottles.

**LDPE** (Low-Density Polyethylene) : Often found in plastic bags and packing films.

**PP** (Polypropylene): Used in bottles caps, food containers and straws.

**HDPE** (High-Density Polyethylene) It is a strong, durable and widely used type of plastic made from petroleum.

**PS** (Polystyrene): Found in disposable cutlery and foam packing.

**PVC** (Polyvinyl Chloride): Less common, but used in pipes and some packing.

**Total Number of Cleanup Events:**

```{r}
total_events <- sum(cleaned_plastics$num_events, na.rm = TRUE)

cat("Total number of cleanup events recorded:", total_events)
```

### Top Contributing Companies to Plastic Waste:

```{r top_companies_chunk, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)

top_companies <- cleaned_plastics %>%
  group_by(parent_company) %>%
  summarise(total_plastics = sum(grand_total, na.rm = TRUE)) %>%
  arrange(desc(total_plastics)) %>%
  slice_head(n = 10)

top_companies %>%
  kable(
    caption = "Table 1. Top Companies Contributing to Plastic Waste",
    col.names = c("Company", "Total Plastics"),
    align = "lc",  # left-align company, center-align plastics
    format = "html"  # use "latex" for PDF output
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f7f7f7")

```

Table 1 shows the top 10 companies responsible for the most plastic waste, based on the number of plastic items collected. Interestingly, the majority of the waste—over 400,000 items—was unbranded, which means it couldn’t be traced back to a specific company. This highlights a major gap in accountability when it comes to plastic pollution. Among the branded companies, Coca-Cola had the highest count with 23,823 items, followed by La Doo, Nestlé, and Unilever. Overall, the data shows that while big companies do play a major role, a significant amount of plastic waste still comes from unknown and difficult to trace sources. Let's dive deeper into the type of plastic waste generated and which country does so in large masses.

```{r country_plastic_relation, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
#Get top 10 countries by total plastic collected
top_countries <- cleaned_plastics %>%
  group_by(country) %>%
  summarise(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_plastic)) %>%
  slice_head(n = 10) %>%
  pull(country)

#merged Sravan's input here

#Summarize plastic type totals for those countries
plastic_types_by_country <- cleaned_plastics %>%
  filter(country %in% top_countries) %>%
  group_by(country) %>%
  summarise(across(c(hdpe, ldpe, o, pet, pp, ps, pvc), sum, na.rm = TRUE)) %>%
  pivot_longer(-country, names_to = "plastic_type", values_to = "count")

#summary(plastic_types_by_country$count)
skim(plastic_types_by_country$count)


```

right skewed  - use of transform to normalize and visualize log transformation on a skewed dataset is a common and effective form of normalization.

**Types Collected by Top 10 Countries:**

```{r fig-country_plastic_heatmap, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=10}
# Top 10 countries
top_countries <- cleaned_plastics %>%
  group_by(country) %>%
  summarise(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_plastic)) %>%
  slice_head(n = 10) %>%
  pull(country)

# Summarise plastic types
plastic_types_by_country <- cleaned_plastics %>%
  filter(country %in% top_countries) %>%
  group_by(country) %>%
  summarise(across(c(hdpe, ldpe, o, pet, pp, ps, pvc), sum, na.rm = TRUE)) %>%
  pivot_longer(-country, names_to = "plastic_type", values_to = "count")

# Heatmap with labels
ggplot(plastic_types_by_country, aes(x = plastic_type, y = fct_reorder(country, -count), fill = count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = ifelse(count > 0, count, "")), size = 3, color = "white") +
  scale_fill_viridis_c(option = "D", trans = "log1p") +
  labs(
    title = "Plastic Types Collected by Top 10 Countries",
    x = "Plastic Type",
    y = "Country",
    fill = "Items Collected (log scale)",
    caption = "Source: TidyTuesday 2021-01-26"
  ) +
  theme_minimal(base_size = 10)

```

**Plastic Waste by Country and Type what we found**

The heat map represents the top 10 countries with the highest amount of plastic waste collected during cleanup events. It also breaks down which types of plastic were most common in each country. The brighter the color, the more items were found.

-   In the Philippines most of the waste came from Coca-Coal products over 12,000 pieces.

-   In Indonesia Nestle was linked to a large portion of the waste.

-   Pepsi Co was a major contributor in India and Mexico.

**Plastic Collected Trend Over Years:**

```{r}
# Plastics collected by year
yearly_summary <- cleaned_plastics %>%
  group_by(year) %>%
  summarise(total_collected = sum(grand_total, na.rm = TRUE))

ggplot(yearly_summary, aes(x = year, y = total_collected)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point() +
  labs(title = "Plastic Collection Trend Over Years", x = "Year", y = "Total Plastics") +
  theme_minimal()
```

**Top Contributing Companies by countries:**

The below chart shows which company contributed the most plastic waste in each country. Through this it is very helpful in identifying the biggest plastic polluter per country making it clean which companies are causing the most damage in specific regions.

```{r}
top_company_by_country <- cleaned_plastics %>%
  group_by(country, parent_company) %>%
  summarise(total = sum(grand_total, na.rm = TRUE), .groups = "drop") %>%
  group_by(country) %>%
  slice_max(total, n = 1)

ggplot(top_company_by_country, aes(x = fct_reorder(country, total), y = total, fill = parent_company)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Plastic Polluters by Country",
    x = "Country",
    y = "Total Plastic",
    fill = "Company"
  ) +
  theme_minimal(base_size = 10)
```

**Plastic Polluted Density by Country:**

This section lists the top 10 countries where the most sample of plastic waste were collected and it gives an idea of where most of the data was gathered which could be because of more pollution in those places or more people collecting data there. which helps us to understand which countries are most represented in the data set.

```{r}
pollution_density <- cleaned_plastics %>%
  count(country, name = "observations") %>%
  arrange(desc(observations))

pollution_density %>%
  slice_head(n = 10) %>%
  kable(
    caption = "Top 10 Countries by Number of Observations",
    col.names = c("Country", "Samples Collected")
  ) %>%
  kable_styling()
```

**Plastic Types Shared by Top Companies:**

The bar chart breakdown the types contributed by the top 5 most polluting companies to known which kind of plastic are mostly produced by which company. For example one company might be producing more PET bottles while another might contribute more plastic bags.

```{r}
top_companies <- cleaned_plastics %>%
  group_by(parent_company) %>%
  summarise(total = sum(grand_total, na.rm = TRUE)) %>%
  arrange(desc(total)) %>%
  slice_head(n = 5) %>%
  pull(parent_company)

company_type_share <- cleaned_plastics %>%
  filter(parent_company %in% top_companies) %>%
  group_by(parent_company) %>%
  summarise(across(c(hdpe, ldpe, o, pet, pp, ps, pvc), sum, na.rm = TRUE)) %>%
  pivot_longer(-parent_company, names_to = "plastic_type", values_to = "count")

ggplot(company_type_share, aes(x = plastic_type, y = count, fill = parent_company)) +
  geom_col(position = "dodge") +
  labs(
    title = "Plastic Type Distribution for Top 5 Polluting Companies",
    x = "Plastic Type",
    y = "Count",
    fill = "Company"
  ) +
  theme_minimal(base_size = 10)
```

This project investigates the global issue of ocean plastic pollution using data from the 2021 TidyTuesday plastics dataset, originally collected by Break Free From Plastic. The dataset records plastic waste collected during brand audits across multiple countries, identifying both the types of plastic and the companies responsible for producing them.

Our team focused on analyzing contributions to plastic pollution by company, country, and over time. The data was cleaned and standardized to ensure consistent names for companies and plastic types, and non-informative records (like NAs or zero values) were removed to improve data quality.

**Key Analyses Conducted**: Top Polluting Companies: Coca-Cola, Nestlé, and Unilever were found to be the leading contributors to global plastic pollution. These multinational corporations were consistently identified across multiple regions.

**Plastic Types:** PET (Polyethylene Terephthalate) and HDPE (High-Density Polyethylene) emerged as the most commonly found plastic types, often used in bottles and packaging.

**Country Contributions:** Countries like the Philippines, Indonesia, India, and the United States had the highest volumes of collected plastic waste. These patterns may reflect both high consumption and active cleanup participation.

**Trends Over Time:** We analyzed plastic collection by year to identify trends. While collection levels vary, any upward or downward patterns can help assess the impact of policy, awareness, and cleanup programs over time.

**Visuals and Tables:**

Bar charts and summary tables were used to visualize company-wise and plastic-type-wise contributions.

A yearly trend chart was added to highlight plastic waste changes across years.

**Discussion & Recommendations:**

Our findings show that corporate accountability and government policies play a major role in addressing plastic pollution.

**We recommend:**

Stronger enforcement of Extended Producer Responsibility (EPR) policies.

Increased support for brand audits and cleanup efforts.

Encouraging companies to shift toward sustainable packaging materials.

Governments should enforce regulations more strictly to reduce plastic pollution and hold corporations accountable.

**References**

TidyTuesday. (2021). Plastics dataset. Retrieved from

<https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-01-26>

Break Free From Plastic. (2021). Brand audit report. <https://www.breakfreefromplastic.org/>.
